---
title: "CS5801 Resit Coursework Supporting analysis"
author: "2439515"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document: default
version: 1
---

# 0. Instructions 

```{r}
# Add code here to load all the required libraries with `library()`.  
# Do not include any `install.package()` for any required packages in this rmd file.
library(tidyverse)
library(mice)
```

# 1. Organise and clean the data

## 1.1 Subset the data into the specific dataset allocated
 
```{r}
# Only change the value for SID 
# Assign your student id into the variable SID, for example:
SID <- 2439515               # This is an example, replace 2101234 with your actual ID
SIDoffset <- (SID %% 50) + 1    # Your SID mod 50 + 1

load("property_data.RDa")
# Now subset the student data set
# Pick every 50th observation starting from your offset
# Put into your data frame named mydf (you can rename it)
mydf <- property_data[seq(from=SIDoffset,to=nrow(property_data),by=50),]
```

## 1.2 Data quality plan

```{r}
head(mydf) # View first 6 rows of the data set
```

```{r}
tail(mydf) # View last 6 rows of the data set
```

```{r}
summary(mydf) # Summary of the data set
```
```{r}
str(mydf) # Structure of the data set
```

```{r}
sum(is.na(mydf)) # Check for any NAs in the data set
```

```{r}
unique(mydf$rental_ind) # Unique function to see for any duplicates
```

```{r}
unique(mydf$rented) # Unique function to see for any duplicates
```

```{r}
unique(mydf$parking) # Unique function to see for any duplicates
```

```{r}
unique(mydf$property_id) # Unique function to see for any duplicates
```

```{r}
unique(mydf$type) # Unique function to see for any duplicates
```

```{r}
table(mydf$square_footage) # Table function to show how many instances in each square footage
```

```{r}
table(mydf$type) # Table function to see how many instances of wrongly spelled 'Dulpex'
```

## 1.3 Data quality analysis findings

### Issue 1

```{r}
print(mydf[mydf$square_footage < 100, ]) # Check for any houses under 100 square foot 
```

### Issue 2

```{r}
unique(mydf$type) # Check the for list of property types and view misspellings or duplicates
```

### Issue 3

```{r}
print(mydf[mydf$property_age < 0, ]) # View any properties that have a negative value in property age
```

## 1.4 Data cleaning  
 
### Addressing Issue 1

```{r}
# Using MICE package to impute when square footage under 100
mydf_subset <- mydf %>% 
  select(-property_id) %>% # Remove property ID to avoid MICE issues
  mutate(across(where(is.character), as.factor)) %>%  # Convert to factors to avoid MICE issues
  mutate(square_footage = ifelse(square_footage < 100, NA, square_footage)) # Replace with NA

# Add imputed subset to a data frame
imputed <- mice(mydf_subset, seed = 2025) 
mydf_complete <- complete(imputed)

# Add back the property_id
mydf_complete$property_id <- mydf$property_id

# Add back to original data frame 
mydf <- mydf_complete
```

```{r}
# Print the total of houses with square footage under 100 to check
print(sum(mydf[mydf$square_footage < 100, ]))
```

### Addressing Issue 2

```{r}
# To fix the spelling mistake 
mydf <- mydf %>%
              mutate(type = recode(type,
                                                      "Dulpex" = "Duplex",
                                                      )) 
```

```{r}
unique(mydf$type) # Check to see its fixed and there is no misspellings of property types
```

### Addressing Issue 3

```{r}
# Fix to make property age values into absolute numbers 
mydf <- mydf %>%
              mutate(property_age = abs(property_age))
```

```{r}
# Check to see if fixed and there are any properties with age below 0 
print(sum(mydf[mydf$property_age < 0, ])) 

```

# 2. Exploratory Data Analysis (EDA)

### Element 1

```{r}
# GGPlot scatter plot for Sale price vs Square footage
ggplot(data = mydf, aes(y = sale_price, x = square_footage)) +
  geom_point() + geom_smooth(method = 'lm') +
  ggtitle("Sale Price VS Square footage") +
  labs(y = "Sale Price", x = "Square footage of property") 
```

### Element 2

```{r}
# GGplot box plots to compare different categories of property type and their sale prices
ggplot(data = mydf, aes(y = sale_price, x = type)) +
  geom_boxplot() +
  ggtitle("Sale Price VS Property Type") +
  labs(y = "Sale Price", x = "Property Type") 
```

### Element 3

```{r}
# GGplot box plots to compare number of bedrooms and their sale prices
ggplot(data = mydf, aes(y = sale_price, x = factor(bedrooms))) +
  geom_boxplot() +
  ggtitle("Sale Price VS Number of Bedrooms") +
  labs(y = "Sale Price", x = "Bedrooms") 
```

# 3. Modelling

### Proposed model
```{r}
# ANCOVA model of sale price against bedrooms, square footage and bedroom type
sales_aov <- (aov(mydf$sale_price~ mydf$bedrooms + mydf$square_footage + mydf$type)) 

summary.lm(sales_aov)
```

### Proposed model diagnostics
```{r}
# plotting each individually against sale price

# linear plotting square footage against sale price
plot(lm(mydf$sale_price~ mydf$square_footage))

# linear plotting property type against sale price
plot(lm(mydf$sale_price~ mydf$type))

# linear plotting bedrooms against sale price
plot(lm(mydf$sale_price~ mydf$bedrooms))
```

# 4. Modelling another dependent variable

### EDA for model

```{r}
# Change rental_ind into a factor to avoid issues 
mydf <- mydf %>% 
  mutate(rental_ind = factor(rental_ind, 
                             levels = c(0, 1),      # 0 first, 1 second
                             labels = c("Not rented", "Rented")))
```

```{r}
ggplot(data = mydf, aes(y = sale_price, x = rental_ind)) +
  geom_boxplot() +
  ggtitle("Rental Indicator VS Sale Price ") +
  labs(y = "Sale Price", x = "Rental Indicator") 
```

```{r}
ggplot(data = mydf, aes(y = square_footage, x = rental_ind)) +
  geom_boxplot() +
  ggtitle("Rental Indicator VS Sqaure Footage ") +
  labs(y = "Square Footage", x = "Rental Indicator") 
```

```{r}
ggplot(data = mydf, aes(y = distance_km, x = rental_ind)) +
  geom_boxplot() +
  ggtitle("Rental Indicator VS Distance from city ") +
  labs(y = "Distance from the city", x = "Rental Indicator") 
```

```{r}
ggplot(data = mydf, aes(y = property_age, x = rental_ind)) +
  geom_boxplot() +
  ggtitle("Property age VS Rental Indicator ") +
  labs(y = "Property Age", x = "Rental Indicator") 
```

### Proposed Model

```{r}
# Building the Logistic Regression model
mydf_glm <- glm(rental_ind ~ square_footage + distance_km + property_age,
              data   = mydf,
              family = binomial)
summary(mydf_glm)
```

### Model intepretation
```{r}
# Step wise function on the model 
step(mydf_glm)

# Coefficient values to find the odds ratios 
exp(coef(mydf_glm))
```

# References  

OpenAI (2025). ChatGPT (July 2025 version). OpenAI. Available at: https://openai.com/chatgpt 
[ChatGPT was used for some syntax and code assistance for example when using the mice package or changing property age into absolute number, changing rental indicator into a factor and other parts of code.]

tidyverse (2023). Wickham,H., Averick,M., Bryan,J., Chang,W., McGowan,L.D.A. and Yutani,H. (2023) tidyverse: Easily Install and Load the “Tidyverse” (R package, version 2.0.0). Vienna: R Foundation for Statistical Computing.

Mice (2011). van Buuren, S. and Groothuis‑Oudshoorn,K. (2011) ‘mice: Multivariate Imputation by Chained Equations in R’, Journal of Statistical Software, 45(3), pp.1–67. R package version 3.16.0.